---
title: "Bellabeat_CS"
author: "Diego Barón"
date: "2024-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## First load the necessary libraries
```{r Loading libraries}
library("tidyverse")
library("ggplot2")
library("skimr")
library("janitor")
library("dplyr")
library("here")
```
## Now the import of the datasets
For my case of study I decided to focus on the Time watch which tracks activity, sleep and stress, is worth to make clear that according to some studies from medical papers the stress can be measure by the heart rate, that's not the only factor that contributes to the stress but is a critical point to detect it and that's why I will be using the Activity data set, the sleep data set and the hearth rate data set.
```{r loading datasets}
activity <- read_csv("dailyActivity_merged.csv")
sleep <- read_csv("minuteSleep_merged.csv")
stress <- read_csv("heartrate_seconds_merged.csv")
```
## Preparation of the data sets
First lets see some statics of our data sets:
```{r Summary of the data}
summary(activity)
summary(sleep)
summary(stress)
```
The first thing to notice is that the date in all data sets are in MM/DD/YYYY format also they are not recognized as dates but character type of data and on the sleep and stress data sets the date is long date meaning it also have a time stamp, so lets start by formatting our data.
### More data statics
Now lets check the data sets for unique participants
```{r Counting unique users}
n_distinct(activity$Id)
n_distinct(sleep$Id)
n_distinct(stress$Id)
```
With this new set of statics we can see that:

* The Activity data set have 35 participants
* The Sleep data set have 23 participants
* The Stress data set have 14 participants

## Data Formatting
Now with this first statics let's start with the formatting of the date field in every data set:
```{r Formatting dates}
activity$ActivityDate=as.POSIXct(activity$ActivityDate, format="%m/%d/%Y", tz=Sys.timezone())
sleep$date=as.POSIXct(sleep$date, format="%m/%d/%Y %I:%M:%S", tz="")
sleep$shortDate <- format(sleep$date, format = "%m/%d/%y")
stress$Time=as.POSIXct(stress$Time, format="%m/%d/%Y %I:%M:%S", tz="")
stress$shortDate <- format(stress$Time, format = "%m/%d/%y")
```
With this I completed the formatting of the dates as well to include an extra separate column in sleep and stress data set for the time and date, this is important later on for analysis purposes.

## More statistics
Now lets make some summaries for data in each data set:

### Activity data set
```{r Activity summary A}
activity %>% 
  select(TotalSteps,TotalDistance,SedentaryMinutes,Calories) %>% summary()
```
```{r Activity summary B}
activity %>% 
  select(VeryActiveMinutes,FairlyActiveMinutes,LightlyActiveMinutes) %>% summary()
```
This first selective summaries show us some insights:

*The mean total steps are 6547
*The mean total distance traveled is 4.664 miles
*The mean sedentary time in minutes are 995.3 mins (16.58 hrs aprox.)
*The mean of very active (intense) activity minutes is 16.62 mins
*The mean of fairly active activity is 13.07 mins
*The mean of light activity in minutes is 170.1 mins (2.8 hrs aprox.)
*This means that most of the steps and distance traveled is during light activity

### Sleep data set
```{r Sleep summary A}
sleepDF <- sleep %>% 
  group_by(Id) %>% 
  summarize(totalMinutes = sum(value),
            averageHours = totalMinutes/60 ,
            MeasuredDays = n_distinct(shortDate),
            bedTimeDay = averageHours/MeasuredDays)
```
This filter let us see some more insights of the sleep data such as:

*Not all participants used the sleep function the same amount of days
*There is a high chance that the data was not collected equally due to technical fails

Further I will return to this summary to get other insights such as *average bed time per day*, *average days of usage*
and *average hours slept*, for now lets see a similar summary but for the heart rate data set.

### Heart rate (Stress) data set
```{r Stress summary A}
stressDF <- stress %>% 
  group_by(Id) %>% 
  summarize(MeasuredDays = n_distinct(shortDate), averageHeartRateDay = sum(Value)/MeasuredDays)
```
With this we can extract the next insights:

*The American Heart Association have a definition of normal heart rate for adults in 60 - 100 BPM (Beats Per Minute)
*Which means that a normal heart rate per day should be around 86,400 - 144000 BPMs per day
*The results above suggest that most of the participants in the heart rate data set where under some kind of extreme stress, i.e physical exercise or stressful situations that accelerate their heart rate.

Further I will combine the data sets in order prove two hypothesis:

*The increase in the heart rate is due to high activity levels
*The most of the participants of the heart rate data set also have the most sleep hours

## Activity and Stress Merge
```{r Merging Activity and stress data sets}
activityMerged <- merge(stressDF,activity, by = c('Id'))
head(mergedData)
```

## Sleep and Stress Merge
```{r Merging Activity and stress data sets}
mergedData <- merge(stressDF,sleepDF, by = c('Id'))
head(mergedData)
```
## Making visualitation

With the merged data lets create a visualization

```{r calories vs steps, echo=FALSE}
ggplot(data=activity, aes(x = TotalSteps, y = Calories)) +
  geom_point(color = "purple") + geom_smooth() + labs(title = "Total steps vs calories")
```
As we can se from this graph the calories burned are not always related to steps done, as we can see the calories burned
tend to concentrate between 0 to 10,000 steps.
```{r Heart rate vs Active minutes, echo=FALSE}
ggplot(data=activityMerged, aes(x = VeryActiveMinutes, y = averageHeartRateDay)) +
  geom_point() + labs(title = "Active minutes vs Heart rate")
```
On the first hypothesis we can see that the data supports it, in the graph is shown that an elevated heart rate is ocurring
in the range of 0 to 100 minutes of high activity.

```{r Heart rate vs Bed time, echo=FALSE}
ggplot(data=mergedData, aes(x = bedTimeDay, y = averageHeartRateDay)) +
  geom_point() + labs(title = "Heart rate vs Bed time")
```
Unfortunately the second hypothesis is inconclusive due to the lack of data, there is so little cases of study to detect
a trend or pattern.

## Conclusion
